---
title: "Analysis 10X dataset"
output: 
  html_document:
    fig_caption: yes
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
    theme: cerulean
editor_options: 
  chunk_output_type: console
params:
  prefix: "FH_CBRA-F210DCRIY"
  resDEG: "SCT_snn_res.0.25"
  percent.mito: 0.15
  regress.batch: FALSE
  sc.transform: TRUE
---



```{r setup, include=FALSE, cache=FALSE}
library(future)
plan("multiprocess", workers = 2)
options(future.globals.maxSize = 8000 * 1024^2)

ifelse(dir.exists("/mntp/scc/project_workspace/"),
       cremwd <- "/mnt/scc/",
       cremwd <- "/restricted/projectnb/crem-bioinfo/")
master_path <- file.path(cremwd, "/project_workspace/20_01_14_marally/calculations/analysis/")
prefix <- params$prefix
calculations_path <- paste0(master_path, prefix, "/")
plots_path <- file.path(calculations_path, "plots/")
rds_path <- file.path(calculations_path, "rds/")
cache_path <- file.path(calculations_path, "cache/") # it must end in a slash
dir.create(rds_path, recursive = T)
dir.create(plots_path, recursive = T)

knitr::opts_chunk$set(echo = FALSE, message = FALSE,error = FALSE,warning = FALSE, cache = TRUE, cache.path = cache_path, fig.width=9,fig.height=7, autodep=TRUE, collapse = FALSE, fig.path = plots_path, fig.keep = "all", comment = "##", dev = c("png", "pdf"), cache.lazy = FALSE)
options(tibble.print_max = 150)

library(tidyverse)
library(Seurat)
library(Matrix)
library(matrixStats) #rowVars
library(magrittr)
library(dplyr)
library(RColorBrewer)
library(kableExtra)
library(data.table) # getting counts
library(knitr)
library(clustree)
```


```{r load_objects, eval=FALSE}
sc.pre <- readRDS(paste0(rds_path, "sc.prefilt.Rds"))
sc <- readRDS(paste0(rds_path, "sc.Rds"))
```


# QC filtering

```{r load_sc_qc, fig.height=4}
sc.counts <-  Read10X(paste0(cremwd, "/project_workspace/20_01_14_marally/calculations/cellranger_count/", prefix, "/outs/filtered_feature_bc_matrix"))
sc.pre <- CreateSeuratObject(counts = sc.counts, min.cells = 5, min.features = 0, project = prefix)
cat("UMI counts per cell")
summary(apply(GetAssayData(sc.pre, slot = "counts"), 2, sum))
cat("UMI counts per gene")
summary(apply(GetAssayData(sc.pre, slot = "counts"), 1, sum))
sc.pre[["percent.mito"]] <- PercentageFeatureSet(sc.pre, pattern = "^MT-") # range 0 to 100
saveRDS(sc.pre, paste0(rds_path, "sc.pre.Rds"))
cat("Before filtering")
plots <- VlnPlot(sc.pre, features = c("nFeature_RNA", "nCount_RNA", "percent.mito"), ncol = 3, pt.size = 0.1, combine=F)
plots <- lapply(X = plots, FUN = function(p) {
  blank <- element_blank()
  p + theme(axis.line.x = blank, axis.line.y = blank,
            axis.text.x = blank, # comment on multisample analyses
            axis.ticks.x = blank,
            axis.title.x = blank, axis.title.y = blank)})
CombinePlots(plots = plots, legend = "none", ncol = 3)
ggsave(paste0(plots_path, "vln.qc.pre.pdf"))
sc.pre
quantile.threshold <- (100 - ((dim(sc.pre)[2]/2)/1000)) / 100  # real cells are much less than detected barcodes. To estimate the duplet threshold more realistically, divide the barcode number by half

nGene.upper <- quantile(sc.pre$nFeature_RNA, probs = quantile.threshold)
nUMI.upper <- quantile(sc.pre$nCount_RNA, probs = quantile.threshold)
# nGene.upper <- 3500
# nUMI.upper <- 15000
cat("Doublet nGene threshold: ", nGene.upper)
cat("Doublet nUMI threshold: ", nUMI.upper)
keep.cells <- sc.pre[[]] %>% rownames_to_column() %>% dplyr::filter(nFeature_RNA < nGene.upper & nFeature_RNA > 800 & nCount_RNA < nUMI.upper & percent.mito < params$percent.mito) %>% pull(rowname)
sc <- sc.pre[, keep.cells]
saveRDS(keep.cells, paste0(rds_path, "keep.cells.Rds"))

cat("After filtering doublets and degraded cells")
plots <- VlnPlot(sc, features = c("nFeature_RNA", "nCount_RNA", "percent.mito"), ncol = 3, pt.size = 0.1, combine = F) 
plots <- lapply(X = plots, FUN = function(p) {
  blank <- element_blank()
  p + theme(axis.line.x = blank, axis.line.y = blank,
            axis.text.x = blank, # comment on multisample analyses
            axis.ticks.x = blank,
            axis.title.x = blank, axis.title.y = blank)})
CombinePlots(plots = plots, legend = "none", ncol = 3)
ggsave(paste0(plots_path, "vln.qc.after.pdf"))

sc
```


```{r interesting_genes}
interesting_genes <- c(
  "NKX2-1", "EGFP_EXTRA", "TDTOMATO_EXTRA_IVS", "SFTA3", "SFTPB", "SOX2",
  "AFP", "ALB",
  "TFF1", # gut/gastric
  "CDX2", # intestinal
  "SCGB3A2", # type 2 cells
  "TOP2A", "AURKB", "MKI67", "BIRC5", # cell cycle
  "TP63", "KRT5",
  "SCGB1A1", 
  "SFTPC", "TFF2",
  "SOX9", 
  "HP",
  "FOXJ1",
  "BMP3", "FOXP2", "CPM", "FGF9",
  "NKD1",
  "GIF", "CLDN18", "MUC6",
  "TF", "APOA1", "CYP4F3", "CYP3A7",
  "PAX9", "KRT14",
  "SFTA2", "AQP4",
  "AXIN2",
  "CFTR")

cat("Pre-selected genes: ", interesting_genes)
interesting_genes <- interesting_genes[interesting_genes %in% rownames(sc)]
cat("Pre-selected genes expressed in current dataset: ", interesting_genes)
```

# Dimensionality reduction

Regressing out the confounding effect of cell degradation (by modeling the mitochondrial percentage as a latent variable).

```{r dim_red_pca}
if (params$sc.transform) {
  if (params$regress.batch) {
    sc <- SCTransform(sc, vars.to.regress="percent.mito", batch_var = "orig.ident", verbose = FALSE)
  } else {
    sc <- SCTransform(sc, vars.to.regress="percent.mito", verbose = FALSE)
  }
} else {
  sc <- NormalizeData(sc)
  sc <- FindVariableFeatures(sc)
  cat("Variable gene selection: ", length(VariableFeatures(sc)))
  all.genes <- rownames(sc)
  if (params$regress.batch) {
    sc <- ScaleData(sc, features = all.genes, vars.to.regress=c("percent.mito", "orig.ident"), display.progress = F) # default: only scales var.genes, which misses some in heatmap
  } else {
    sc <- ScaleData(sc, features = all.genes, vars.to.regress="percent.mito", display.progress = F) # default: only scales var.genes, which misses some in heatmap
  }
}
 
sc <- RunPCA(sc, features = VariableFeatures(sc), verbose = F)
sc <- ProjectDim(sc, verbose = F)
print(x = sc[['pca']], dims = 1:5, nfeatures = 5, projected = FALSE)

ElbowPlot(sc)
ggsave(paste0(plots_path, "elbow.pdf"))
DimPlot(sc, reduction = "pca")
```

These are the genes driving the first 9 principal components

```{r dim_heatmap, fig.height=11}
DimHeatmap(sc, dims = 1:9, cells = 500, balanced = TRUE)
```


# Clustering

Use UMAP for non-linear dimensionality reduction

Use Louvain algorithm for community detection.

```{r cluster_and_rdim, eval=TRUE, results=FALSE}
sc <- FindNeighbors(sc, dims = 1:20)
sc <- FindClusters(sc, resolution = seq(1.5, 0.25, -0.25)) # reverse order, seurat_clusters takes last run's value
sc <- FindClusters(sc, resolution = seq(0.20, 0.05, -0.05)) # reverse order, seurat_clusters takes last run's value

sc <- RunUMAP(sc, dims = 1:20)
# Idents(sc) <- "orig.ident"
```

Non-linear dimensionality reduction with UMAP

```{r plot_rdim}
DimPlot(sc, reduction = "umap", pt.size = 2) + theme_void() + labs(title ="UMAP")
ggsave(paste0(plots_path, "rdim.pdf"), width = 5, height = 5)
```

Number of cells per cluster at different Louvain resolutions:

```{r}
if (params$sc.transform) {
  table(sc$SCT_snn_res.1.5)
  table(sc$SCT_snn_res.1.25)
  table(sc$SCT_snn_res.0.75)
  table(sc$SCT_snn_res.0.5)
  table(sc$SCT_snn_res.0.25)
  clustree(sc, prefix = "SCT_snn_res.")
} else {
  table(sc$RNA_snn_res.1.5)
  table(sc$RNA_snn_res.1.25)
  table(sc$RNA_snn_res.0.75)
  table(sc$RNA_snn_res.0.5)
  table(sc$RNA_snn_res.0.25)
  clustree(sc, prefix = "RNA_snn_res.")
}

```

# Cell-cycle stage classification

Score G1, G2/M and S signatures in each cell.

Classify phase based on that score.

```{r CC_regression, eval=TRUE}
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
sc <- CellCycleScoring(sc, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)
# head(sc[[]])
grouping <- "Phase"; DimPlot(sc, reduction.use = "umap", group.by =  grouping, pt.size = 2) + NoAxes()

if (params$sc.transform) {
  kable(table(sc$Phase, sc$SCT_snn_res.0.25), caption="Cell cycle stage classification per cluster") %>%   kable_styling() %>%   scroll_box(width = "100%", height = "200px")
} else {
  kable(table(sc$Phase, sc$RNA_snn_res.0.25), caption="Cell cycle stage classification per cluster") %>%   kable_styling() %>%   scroll_box(width = "100%", height = "200px")
}
```

# Molecular Signatures scoring

Compute enrichment for molecular signatures using method from Tirosh et al, Science (2016)

```{r mol_sig_score, fig.height=11}
msig <- scan(paste0(cremwd, "/reference_data/gene_sets/lung.matur.differ.MSig.txt"), what="", sep="\n") # Read in the data
msig <- sapply(msig, function(k) gsub(",$", "", k))
msig <- strsplit(msig, ",[[:space:]]+")  # Separate elements by a comma and one or more whitepace
names(msig) <- sapply(msig, `[[`, 1)  # subset the first vector element and set it as the list element name
msig <- lapply(msig, `[`, -1) # Remove the first vector element from each list element
msig2 <- read_csv(paste0(cremwd, "/reference_data/gene_sets/lung.cancer.msig.csv"), comment = "#", col_names = F)
msig2 <- split(msig2$X1, msig2$X2)
msig <- c(msig, msig2)
sc <- AddModuleScore(sc, features = msig, name = names(msig))
sc <- AddMetaData(sc, sc[[paste0(names(msig), 1:length(names(msig)))]], col.name = names(msig))
for(i in paste0(names(msig), 1:length(names(msig)))) {
  sc[[i]] <- NULL
}
```

# UMAP clusters

```{r facetted_louvain_clusters_UMAP, fig.height=6}
if (params$sc.transform) {
  grouping <- "SCT_snn_res.0.1"; p00 <- DimPlot(sc, group.by =  grouping, plot.title= grouping, pt.size = 2) + theme_void() + labs(title = grouping) 
  grouping <- "SCT_snn_res.0.15"; p01 <- DimPlot(sc, group.by =  grouping, plot.title= grouping, pt.size = 2) + theme_void() + labs(title = grouping) 
  grouping <- "SCT_snn_res.0.2"; p02 <-  DimPlot(sc, group.by =  grouping, plot.title= grouping, pt.size = 2) + theme_void() + labs(title = grouping)
  grouping <- "SCT_snn_res.0.25"; p1 <- DimPlot(sc, group.by =  grouping, plot.title= grouping, pt.size = 2) + theme_void() + labs(title = grouping) 
  grouping <- "SCT_snn_res.0.5"; p2 <-  DimPlot(sc, group.by =  grouping, plot.title= grouping, pt.size = 2) + theme_void() + labs(title = grouping)
  grouping <- "SCT_snn_res.0.75"; p3 <- DimPlot(sc, group.by =  grouping, plot.title= grouping, pt.size = 2) + theme_void() + labs(title = grouping)
} else {
  grouping <- "RNA_snn_res.0.1"; p00 <- DimPlot(sc, group.by =  grouping, plot.title= grouping, pt.size = 2) + theme_void() + labs(title = grouping) 
  grouping <- "RNA_snn_res.0.15"; p01 <- DimPlot(sc, group.by =  grouping, plot.title= grouping, pt.size = 2) + theme_void() + labs(title = grouping) 
  grouping <- "RNA_snn_res.0.2"; p02 <-  DimPlot(sc, group.by =  grouping, plot.title= grouping, pt.size = 2) + theme_void() + labs(title = grouping)
  grouping <- "RNA_snn_res.0.25"; p1 <- DimPlot(sc, group.by =  grouping, plot.title= grouping, pt.size = 2) + theme_void() + labs(title = grouping) 
  grouping <- "RNA_snn_res.0.5"; p2 <-  DimPlot(sc, group.by =  grouping, plot.title= grouping, pt.size = 2) + theme_void() + labs(title = grouping)
  grouping <- "RNA_snn_res.0.75"; p3 <- DimPlot(sc, group.by =  grouping, plot.title= grouping, pt.size = 2) + theme_void() + labs(title = grouping)
}
# grid.arrange(p1, p2, p3, p4, p5, p6, ncol=3)
CombinePlots(plots = list(p00, p01, p02, p1, p2, p3))
ggsave(paste0(plots_path, "rdim.clust.pdf"), width = 10, height = 7)

```


# UMAP pre-selected features

```{r echo=FALSE}
# 4 columns, 9 width, height per line == 9 / 4 == 2.25
x <- interesting_feats <- c("nFeature_RNA", "nCount_RNA", "percent.mito", paste0("PC_", 1:10), names(msig), interesting_genes)
len <- (length(unique(x))/4)*2.25
```

```{r pre_selected_marker_genes_UMAP,  fig.height=len}
interesting_feats <- c("nFeature_RNA", "nCount_RNA", "percent.mito", paste0("PC_", 1:10), names(msig), interesting_genes)
plots <- FeaturePlot(sc, features = interesting_feats, cols = c("#f0f0f0", brewer.pal(9,"OrRd")), pt.size = 1, combine=F)
plots <- lapply(X = plots, FUN = function(p) {
  blank <- element_blank()
  p + theme(axis.line.x = blank, axis.line.y = blank,
            axis.text.x = blank, axis.text.y = blank, 
            axis.title.x = blank, axis.title.y = blank,
            axis.ticks.x = blank, axis.ticks.y = blank)})
CombinePlots(plots = plots, legend = "bottom", ncol = 4)
```


# Differential expression

Differential expression model: MAST, Finak et al.: 
https://genomebiology.biomedcentral.com/articles/10.1186/s13059-015-0844-5 

For a recent comparison on the performance of different methods for single-cell differential expression, see:
https://www.nature.com/articles/nmeth.4612

 
```{r use_seurat, eval=TRUE, fig.height=15}
fmas <- function(sc, res) {
    Idents(sc) <- res
    file_out <- paste0(calculations_path, res, ".DEG.clusterMarkers.xlsx")
    seurat.markers <- FindAllMarkers(sc, test.use = "MAST", only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, verbose = F)
    seurat.markers.summary <- as.data.frame(seurat.markers %>% group_by(cluster) %>% dplyr::filter(p_val_adj < 0.05) %>% top_n(20, abs(avg_logFC)))
    openxlsx::write.xlsx(seurat.markers, file_out, zoom = 100, asTable = TRUE, tableStyle = "none", rowNames = F)
    return(list(all = seurat.markers, top = seurat.markers.summary))
}
resDEG <- params$resDEG
sc.markers.mas.res <- fmas(sc, resDEG)
saveRDS(sc.markers.mas.res, paste0(rds_path, "sc.markers.mas.", resDEG, ".Rds"))
```

Table of top 20 DEG per cluster

The clustering resolution used for differential expression is **`r resDEG`**

```{r}
sc.markers.mas.res[[2]] %>% dplyr::mutate(cluster = cell_spec(
    cluster, color = "white", bold = T,
    background = spec_color(as.numeric(factor(cluster)), end = 0.9, direction = -1)
  )) %>% kable(escape = F, caption=paste0("Top DEG for clusters at ", resDEG )) %>%   kable_styling() %>%   scroll_box(width = "100%", height = "600px")
```


# Heatmap of DEG

Heatmap of top 20 DEG per cluster

The clustering resolution used for differential expression is **`r resDEG`**


```{r heatmap_DEG, fig.height=15}
Idents(sc) <- resDEG  
# switch to RNA normalized for heatmap visualization
if (params$sc.transform) {
  DefaultAssay(sc) <- "RNA"
  sc <- NormalizeData(sc)
  all.genes <- rownames(sc)
  sc <- ScaleData(sc, features = all.genes, vars.to.regress="percent.mito", display.progress = F) # default: only scales var.genes, which misses some in heatmap
  g <- DoHeatmap(sc, features = sc.markers.mas.res[[2]]$gene) + scale_fill_gradientn(colors = c("blue", "white", "red"))
  print(g) # only need to use print if you aren't in a top-level statement (e.g. in a function, in an if/else block, etc.)
  DefaultAssay(sc) <- "SCT"
} else {
  g <- DoHeatmap(sc, features = sc.markers.mas.res[[2]]$gene) + scale_fill_gradientn(colors = c("blue", "white", "red"))
  print(g)
}
```


# UMAP of DEG

The clustering resolution used for differential expression is **`r resDEG`**


```{r echo=FALSE}
# 6 columns, 9 width, height per line == 9 / 6 == 1.5
x <- sc.markers.mas.res[[2]]$gene
len2 <- (length(unique(x))/6)*1.5
```

```{r feat_DEG_UMAP,  fig.height=len2}
plots <- FeaturePlot(sc, features = sc.markers.mas.res[[2]]$gene, cols = c("#f0f0f0", brewer.pal(9,"OrRd")), pt.size = 1, combine=F) 
plots <- lapply(X = plots, FUN = function(p) {
  blank <- element_blank()
  p + theme(axis.line.x = blank, axis.line.y = blank,
            axis.text.x = blank, axis.text.y = blank, 
            axis.title.x = blank, axis.title.y = blank,
            axis.ticks.x = blank, axis.ticks.y = blank)})
CombinePlots(plots = plots, legend = "bottom", ncol = 6)
```


```{r save_object}
saveRDS(sc, paste0(rds_path, "sc.Rds"))
```

```{r export_spring, eval=TRUE}
export.data <- GetAssayData(sc, slot = "data")
export.genes <- rownames(sc)
write.table(as.matrix(export.data), gzfile(paste0(calculations_path, "spring.data.tsv.gz")), sep = "\t", quote = F, col.names = T, row.names = T )
write.table(export.genes, paste0(calculations_path, "spring.genes.tsv"), sep = "\t", quote = F, col.names = F, row.names = F )
if (params$sc.transform) {
  meta.subset <- sc[[c("Phase", "SCT_snn_res.0.05", "SCT_snn_res.0.1", "SCT_snn_res.0.15", "SCT_snn_res.0.2","SCT_snn_res.0.25", "SCT_snn_res.0.5", "SCT_snn_res.0.75", "SCT_snn_res.1")]]
} else {
  meta.subset <- sc[[c("Phase",  "RNA_snn_res.0.05", "RNA_snn_res.0.1", "RNA_snn_res.0.15", "RNA_snn_res.0.2","RNA_snn_res.0.25", "RNA_snn_res.0.5", "RNA_snn_res.0.75", "RNA_snn_res.1")]]
}
groupings <- t(meta.subset) # select metadata
meta.continuous <- sc[[names(msig)]]
custom.color.tracks <- t(meta.continuous)
write.table(custom.color.tracks, paste0(calculations_path, "spring.custom.color.tracks.csv"), sep = ",", quote = F, col.names = F, row.names = T )
write.table(groupings, paste0(calculations_path, "spring.groupings.csv"), sep = ",", quote = F, col.names = F, row.names = T )

```

